---
title: "Get started with bootpq"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with bootpq}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Why bootpq?

A single rarefaction to even sequencing depth throws away part of the data and
depends on one random draw. The conclusion of a downstream test (PERMANOVA, a
Hill-number group comparison, a variation partitioning) can therefore change
from one rarefaction to the next.

`bootpq` removes this fragility by **repeating the analysis across many
rarefaction permutations** — each with a different random seed — and
summarising the results (mean and quantiles). You get an estimate together with
its variability across rarefactions, instead of a single, possibly atypical,
value.

`bootpq` is part of the [pqverse](https://github.com/adrientaudiere) ecosystem
and builds on [`MiscMetabar`](https://adrientaudiere.github.io/MiscMetabar/).

## Installation

```{r, eval = FALSE}
# install.packages("remotes")
remotes::install_github("adrientaudiere/bootpq")
```

## A worked example

We use `data_fungi_mini`, a small fungal metabarcoding dataset shipped with
`MiscMetabar`. We first drop samples with missing covariates.

```{r setup, message = FALSE}
library(bootpq)
library(MiscMetabar)
data(data_fungi_mini)

data_fungi_woNA <- subset_samples(
  data_fungi_mini,
  !is.na(Time) & !is.na(Height)
)
data_fungi_woNA
```

### PERMANOVA across rarefaction permutations

`adonis_rarperm_pq()` runs `vegan::adonis2()` on each rarefied version of the
object and summarises the results. We keep `nperm` small here so the vignette
builds quickly; in practice use the default (`nperm = 99`) or more.

```{r adonis}
res <- adonis_rarperm_pq(
  data_fungi_woNA,
  "Time*Height",
  na_remove = TRUE,
  nperm = 9,
  progress_bar = FALSE
)

# Mean of the PERMANOVA table across the 9 rarefaction permutations
res$mean
```

`res$quantile_min` and `res$quantile_max` give the corresponding lower and upper
quantiles, so you can see how stable each term is across rarefactions.

### Hill-number group tests

`hill_test_rarperm_pq()` compares Hill numbers (q = 0, 1, 2) between the levels
of a factor across rarefaction permutations. It relies on `ggstatsplot`, so the
chunk below is shown but not run during the vignette build.

```{r hill, eval = FALSE}
hill_test_rarperm_pq(data_fungi_woNA, fact = "Height", nperm = 99)
```

### Variation partitioning

`var_par_rarperm_pq()` repeats a variation partitioning across rarefaction
permutations, summarising how much variation each set of explanatory variables
explains.

```{r varpar, eval = FALSE}
var_par_rarperm_pq(
  data_fungi_woNA,
  list_component = list(
    "Time" = c("Time"),
    "Size" = c("Height", "Diameter")
  ),
  nperm = 99
)
```

## Where to go next

- The [function reference](../reference/index.html) documents every argument,
  including how to control the rarefaction depth (`sample.size`) and the
  quantiles reported (`quantile_prob`).
- `bootpq` works on any `phyloseq` object, so the same workflow applies to your
  own data.
